---
title: "Assignment 5: Data Visualization"
author: "Ye Khaung Oo"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse); library(lubridate); library(here); library(cowplot); library(ggplot2); library(ggthemes);
getwd()

peter_paul_df <- read.csv("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
litter_df <- read.csv("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
class(peter_paul_df$sampledate)
class(litter_df$collectDate)
peter_paul_df$sampledate <- as.Date((peter_paul_df$sampledate), format = "%Y-%m-%d")
litter_df$collectDate <- as.Date((litter_df$collectDate), format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
yko_theme <- theme_economist(base_size = 7) +
  theme(axis.text = element_text(color = "blue"), 
        axis.text.x = element_text(color = "darkblue", size = 8),
        axis.text.y = element_text(color = "darkblue", size = 8),
        plot.title =        element_text(color = "red", size = 9),
        axis.ticks =        element_line(6),
        legend.position = "right",
        legend.direction = "vertical",
        legend.background = element_rect(color='lightyellow'
    ),
    legend.title = element_text(
      color="darkgrey", size = 8
    )
  )

theme_set(yko_theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
peter_paul_df %>% 
  ggplot(aes(x=po4,y=tp_ug, color=lakename)) + 
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(lakename))+
  ylab("Total Phosphrous")+
  xlab("Phosphate") +
  xlim(0,50) +
  ylim(0,150)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
#turning months into factor
peter_paul_df$month_f <- factor(
  peter_paul_df$month, 
  levels=1:12,
  labels=month.abb,
  ) 

# Boxplot for Temperature
box_temp <- ggplot(peter_paul_df, aes(x = month_f, y = temperature_C, fill = lakename,)) +
  geom_boxplot() +
  scale_x_discrete(drop=F) +
  labs(title = "Temperature by Month",
       x = "Month",
       y = "Temperature (°C)") 
box_temp

#boxplot for Phosphrous
box_tp <- ggplot(peter_paul_df, aes(x = month_f, y = tp_ug, fill = lakename,)) +
  geom_boxplot() +
  scale_x_discrete(drop=F) +
  labs(title = "Total Phosphorus by Month",
       x = "Month",
       y = "Total Phosphorus (µg/L)")
box_tp

#boxplot for Nitrogen
box_tn <- ggplot(peter_paul_df, aes(x = month_f, y = tn_ug, fill = lakename,)) +
  geom_boxplot() +
  scale_x_discrete(drop=F) +
  labs(title = "Total Nitrogen by Month",
       x = "Month",
       y = "Total Nitrogen (µg/L)")
box_tn

# Combine the three plots using cowplot
three_box <- plot_grid(
  box_temp, box_tp, box_tn, 
  align = "hv", ncol = 1)

# Remove legends from individual plots
three_box <- three_box + theme(axis.title = element_blank(), legend.position = "none")

# Display the combined plot
three_box

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Phosphrous and Nitrogen seem to rise with rising temperatures in the summer months of July and August. Peter Lake seems to have more nutrients amount in the water. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
needles_df <- litter_df %>% 
  filter(functionalGroup == "Needles")

# Create scatter plot of dry mass by date, colored by NLCD class
ggplot(needles_df, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  labs(title = "Needle Litter Dry Mass by Date",
       x = "Date",
       y = "Dry Mass (g)") +
  theme_excel_new()

#7
ggplot(needles_df, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), ncol = 3) + 
  labs(title = "Needle Litter Dry Mass by Date (Facet)",
       x = "Date",
       y = "Dry Mass (g)") +
  theme_excel_new()

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective since the visualization gives out separate timeline for each category of needle group while Plot 6 only provides a timeline with all three needle groups mixed. Plot 6 is much harder to interpret.