---
title: "Assignment 3: Data Exploration"
author: "Ye Khaung Oo"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#getting working directory
getwd()

#loading packages
library(tidyverse)
library(lubridate)

#importing dataset
Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = T)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = T)
#View(Neonics)
#View(Litter)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are considered systemic pesticides and kill all insects indiscrimately. Insects such as bees and butterflies are crucial for the ecosystem maintaining healthy soil, recycling nutrients and pollinating flowers. Without the insects, our food security is severely threatened. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris help decompose organic materials, sending nutrients back into the food chain. Litter and woody debris are also a source of energy for aquatic ecosystems, and help release carbon dioxide slowly into the atmosphere. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Spatial Sampling Design - executed at terrestrial NEON sites that contain woody vegetation >2m tall.
 2. Temporal Sampling Design - ground traps sampled once every year. Elevated traps varies by vegetation present at site.
 3. Users are advised to leverage data from vegetation structure to provide context to litter data.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}

dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}

summary(Neonics$Effect)

```

> Answer: Most common effects observed are population and mortality.These effects are critical to understand the population change in insect groups since insects have profound ecological significance in terrestrial ecosystems.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}

specs_sum <- summary(Neonics$Species.Common.Name)
sorted_specs_sum <- sort(specs_sum, decreasing = TRUE)
sorted_specs_sum

```

> Answer: Six most commonly studied species are honey bee, parasitic wasp, buff tailed bumblebee, carniolan honey bee and bumble bee. Bees are important pollinators contributing directly to food security. UN Environment Program stated bees are responsible for 75 percent of our food chain. Parasitic wasp helps control agricultural pests.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}

class(Neonics$Conc.1..Author.)

```

> Answer: `Conc.1..Author.` column returns 'factor' class. It is not numeric because of the symbols such as '/' and '<'. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}

ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year),bins=30) +
  scale_x_continuous(limits = c(1980,2023))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}

#ggplot(Neonics, aes(x=Test.Location, fill=Test.Location)) +
  #geom_bar()
  
ggplot(Neonics, aes(x=Publication.Year, color=Test.Location)) +
  geom_freqpoly(bins=20) 
  
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Most common test locations are Lab and Field Natural. Before the early 2000s, it seems like test locations used to be done on the natural fields. The lab environment must have been siginificantly improved with new technology after the early 2000s, and therefore most test locations have shifted to the lab.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}

ggplot(Neonics, aes(x=Endpoint)) +
  geom_bar(fill="darkgreen") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

```

> Answer: The two most common endpoints are NOEL and LOEL. NOEL is short for 'No-observable-effect-level' and LOEL is short for 'lowest-observable-effect-level'. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}

class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate, format="%Y-%m-%d")
class(Litter$collectDate)
Litter$collectDate

unique_dates <- unique(Litter$collectDate)
august_dates <- unique_dates[month(unique_dates) == 8 & year(unique_dates) == 2018]
august_dates

```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

unique_plots <- unique(Litter$plotID)
length(unique_plots)
unique_plots

```

> Answer: Unique function looks for duplicates and eliminates those to return unique values with particular counts. Summary function returns descriptive analysis such as mean, medium, mode, etc.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

ggplot(Litter, aes(x=functionalGroup)) +
  geom_bar(fill="darkblue")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}

ggplot(Litter, aes(x=functionalGroup, y=dryMass)) +
  geom_boxplot(fill="purple")

ggplot(Litter, aes(x=functionalGroup, y=dryMass)) +
  geom_violin(fill="orange")

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The dataset is small and therefore the violin plot is having difficulties showing the data distribution in each category. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles tend to have the highest biomass.
