---
title: "Assignment 10: Data Scraping"
author: "Ye Khaung Oo"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE, tidy=TRUE}
#1 
library("tidyverse"); library("rvest"); library("ggthemes");
library("ggplot2"); library("lubridate"); library("dplyr")
getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website, tidy=TRUE}
#2 
the_website <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data, tidy=TRUE}
#3 
water_system_name <- html_nodes(the_website, 'div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text(trim = TRUE)
pwsid <- html_nodes(the_website, 'td tr:nth-child(1) td:nth-child(5)') %>% html_text(trim = TRUE)
ownership <- html_nodes(the_website, 'div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text(trim = TRUE)
maximum_day_use <- html_nodes(the_website, 'th~ td+ td') %>% html_text(trim = TRUE) %>% as.numeric()
```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, tidy=TRUE}
#4 
water_system_name <- rep("Durham", 12)
pwsid <- rep("03-32-010", 12)
ownership <- rep("Municipality", 12)
months <- c("Jan","May","Sep","Feb","Jun","Oct","Mar","Jul","Nov","Apr","Aug","Dec")
year <- rep("2022",12)
dates <- as.Date(paste("2022", months, "01", sep = "-"), format = "%Y-%b-%d")

df <- data.frame(
  df_water_system_name = water_system_name,
  df_pwsid = pwsid,
  df_ownership = ownership,
  df_maximum_day_use = maximum_day_use,
  df_date = dates
  )

#5 
ggplot(df, aes(x = df_date, y = df_maximum_day_use)) +
  geom_line(group = 1) +
  labs(x = "Month", y = "Maximum Daily Withdrawals") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_clean() +
  ggtitle("Maximum Daily Withdrawals - 2022 Monthly Data")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function, tidy=TRUE}
#6. 
scraping_data <- function(pwsid,year) {
  the_url <- paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=', pwsid, '&year=', year)
  the_website <- read_html(the_url)

#scraping data from website  
water_system_name <- html_nodes(the_website, 'div+ table tr:nth-child(1) td:nth-child(2)') %>% html_text(trim = TRUE)
pwsid <- html_nodes(the_website, 'td tr:nth-child(1) td:nth-child(5)') %>% html_text(trim = TRUE)
ownership <- html_nodes(the_website, 'div+ table tr:nth-child(2) td:nth-child(4)') %>% html_text(trim = TRUE)
maximum_day_use <- html_nodes(the_website, 'th~ td+ td') %>% html_text(trim = TRUE) %>% as.numeric()

#assigning months and mutating into dates
months <- c("Jan","May","Sep","Feb","Jun","Oct","Mar","Jul","Nov","Apr","Aug","Dec")
dates <- as.Date(paste(year, months, "01", sep = "-"), format = "%Y-%b-%d")

#creating dataframe
df <- data.frame(
  df_water_system_name = rep(water_system_name, 12),
  df_pwsid = rep(pwsid,12),
  df_ownership = rep(ownership,12),
  df_maximum_day_use = maximum_day_use,
  df_months = months,
  df_year = rep(year,12),
  df_date = dates
  )
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, tidy=TRUE}
#7 
durham <- scraping_data('03-32-010','2015')

ggplot(durham, aes(x = df_date, y = df_maximum_day_use)) +
  geom_line(group = 1) +
  labs(x = "Month", y = "Maximum Daily Withdrawals") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_clean() +
  ggtitle("Maximum Daily Withdrawals - 2015 Monthly Data") +
  theme(plot.title = element_text(hjust = 0.5))
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, tidy=TRUE}
#8 
asheville <- scraping_data('01-11-010','2015')

two_towns <- bind_rows(durham,asheville)

ggplot(two_towns, aes(x = df_date, y = df_maximum_day_use, color = df_water_system_name)) +
  geom_line() +
  labs(x = "Month", y = "Maximum Daily Withdrawals", color = "Town") +
  theme_clean() +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  ggtitle("Asheville Vs Durham 2015") +
  theme(plot.title = element_text(hjust = 0.5))
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, tidy=TRUE }
#9 
ash_years <- 2010:2021
ash_pwsid <- rep('01-11-010', length(ash_years))

ash_df <- map2(ash_pwsid, ash_years, scraping_data)
ash_df <- bind_rows(ash_df)

ggplot(ash_df, aes(x = df_date, y = df_maximum_day_use)) +
  geom_line() +
  geom_smooth(method = 'loess', se = FALSE, color = 'red') +
  labs(x = "Date", y = "Maximum Daily Withdrawals", color = "Town") +
  theme_clean() +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +
  ggtitle("Maximum Daily Withdrawals 2010-2021 (Asheville)")
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
> Yes, there is an upward trend in Asheville's water usage.
